% !TEX root = ../main.tex
\clearemptydoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgements}

\vspace*{2cm}

\begin{center}
{\Large \textbf{Acknowledgments}}
\end{center}

\vspace{1cm}

Thanks to the Rust community for providing a welcoming space where everyone can
learn about this wonderful language, in particular to Steve Klabnik and the
rest of the Rust team for always having ease to learn into their goals. I would
also like to thank particularly to Edward Burtescu, Ralf Jung and Oliver
Scherer whose explanations and conversations with me were the basis for this
work.

Nicolás Cardozo has my full gratitude for welcoming an unknown physics graduate
who wanted to learn computer science into his office. Thank you for always
having time for my questions and guiding me though this new academic path that
I have chosen. Also thanks to the rest of the FLAGLAB group for keeping the C
alive in DISC.

Thanks to my parents for all the support and love through my life. Thank you
for always putting me first and giving me the best education. Thank you mom for
teaching me how to be an exceptional person by example. Thank you dad for
answering every single one of my 8-year-old questions.

Before ending these words, I want to thank my wife Diva for being the
unconditional partner who always supported me. Thank you for sacrificing a
comfortable life with me for our goals and dreams, and for taking care of
everything so I could finish my Master. Now we have a new challenge ahead, and
I am not scared of taking it by your side. I love you (and Milo does too).
